---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

##Loading and preprocessing the data
- No extra packages loaded, just packages loaded in Rstudio as default used

- Set working directory
```{r}
setwd("/Users/hss1g09/DataScience/ReproducibleResearch")
```
- Make new folder
```{r}
if (!file.exists("Assign1")){
        dir.create("Assign1")
}
```
- Set working directory to new folder
```{r}
setwd("/Users/hss1g09/DataScience/ReproducibleResearch/Assign1")
```
- define URL of data
```{r}
fileURL<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
```
- download
```{r}
download.file(fileURL, destfile="/Users/hss1g09/DataScience/ReproducibleResearch/Assign1/repdata_2Fdata_2Factivity.zip", method="curl")
```
- time stamp
```{r}
dateDownload<-date()  #records when downloaded
dateDownload
```
- unzip and read data into a table: 'tab'
```{r}
data<-unzip("/Users/hss1g09/DataScience/ReproducibleResearch/Assign1/repdata_2Fdata_2Factivity.zip")
tab<-read.csv(data)
```

##How does the data look like? (not in assignment)
```{r}
head(tab)
tail(tab)
str(tab)
summary(tab)
```

##Part 1: What is mean total number of steps taken per day?
###1 - Histogram
- calculate the steps taken each day
```{r}
sum_steps<-tapply(tab$steps,tab$date,sum) 
```
- make the histogram of the total number of steps taken each day
```{r}
hist(sum_steps, breaks=seq(0,25000,2500), xlim=c(0,25000),ylim=c(0,20), xlab=c("Steps"),
     ylab=c("Frequency [Days]"), main=c(), plot=TRUE)
```

###2 - Mean and median 
- both the mean and median can be displayed with summary()
```{r}
summary(sum_steps)
```

##Part 2: What is the average daily activity pattern?
###1 - Time series plot
- calculate the mean of each 5-minute interval over all days
- NA are ignored with na.rm=TRUE
```{r}
mean_steps_interval<-tapply(tab$steps,tab$interval,mean, na.rm=TRUE)
```
- to plot the data the mean_steps_interval are converted into a data frame: mean_int
- the data are then plotted using the basic plotting system
```{r}
mean_int<-data.frame(mean_steps_interval)
mean_int$interval<-unique(tab$interval)

head(mean_int) #quick look at the new table

par(mfrow=c(1,1))
plot(mean_int$mean_steps_interval~mean_int$interval,type="l", xaxt="n", 
     xlab=c("Time of day [h]"), ylab=c("Number of steps"))
axis(1, at=c(0, 400, 800, 1200, 1600, 2000,2355),labels=c("00:00","4:00","8:00","12:00",
                                                    "16:00","20:00","23:55"))
```

###2 - max steps in 5 min interval
- calculate the interval with the most steps taken over all days
```{r}
max_int<-mean_int[which(mean_int$mean_steps_interval== max(mean_int$mean_steps_interval)),]
max_int$interval
```

##Part 3:Imputing missing values
###1 - Total number of NA
- calculate the total number of NA for each column: steps, date, interval
```{r}
steps_NA<-is.na(tab$steps)
sum(steps_NA)
date_NA<-is.na(tab$date)
sum(date_NA)
interval_NA<-is.na(tab$interval)
sum(interval_NA)
```

###2 - Fill NA with mean of 5 min interval
- each NA is replaced with the mean value of the interval over all days and saved as table: new_tab
- within the for loop the column date is convertic to numeric, thus the a column new_date is inserted to the table new_tab after the for loop

```{r}
#tab$new_interval<-tab$interval+1
new_tab<-matrix(data=NA,ncol=dim(tab)[2],nrow=dim(tab)[1])
colnames(new_tab)<-names(tab)
counter<-1
int<-seq(0,2355,5)
for (i in int){
        
        Loop_interval<-tab[which(tab$interval==i),]
        mean_steps<-mean(Loop_interval$steps,na.rm=TRUE)
        
        
        for (d in Loop_interval$date){
                
                Loop_date<-Loop_interval[which(Loop_interval$date==d),]
                
                if(is.na(Loop_date$steps)==TRUE) 
                        {new_tab[counter, 1]<-mean_steps
                }else{
                        new_tab[counter,1]<-Loop_date$steps}
                       
              
                new_tab[counter,2]<-Loop_date$date
                new_tab[counter,3]<-i
               
                
                counter<-counter +1   
        }
       
}
```

- convert matrix to data.frame
```{r}
new_tab_df<-data.frame(new_tab)
```
- differently sorted version of the original table in oder of the intervals to match new_tab order
```{r}
tab_interval_sorted<-tab[order(tab$interval),]
```

###3 - New data table
- the final new data table contains the extra column new_date 
```{r}
new_tab_df$new_date<-tab_interval_sorted$date

new_tab_df<-new_tab_df[order(new_tab_df$new_date),]
```
comparison tab and new_tab
```{r}
tab[1:20,]
new_tab_df[1:20,]
```

###4 - Histogram
- calculate the steps taken each day when NA filled in
```{r}
sum_steps_filledNA<-tapply(new_tab_df$steps,new_tab_df$new_date,sum) 
```
- make histogramm and display mean and median with summary()
```{r}
hist(sum_steps_filledNA,breaks=seq(0,25000,2500), xlim=c(0,25000),ylim=c(0,30), xlab=c("Steps"),
     ylab=c("Frequency [Days]"), main=c(), plot=TRUE)

summary(sum_steps_filledNA)
```

##Part 3: Are there differences in activity patterns between weekdays and weekends?
- add a new column "weekday" to new_tab (with replaced NA) with the corresponding weekday to each date
- split tab into weekend end: weekday == Saturday or Sunday and week != Saturday or Sunday
- add a new comlum "day_type" to each new table
- merge tables together again, to have a table with both week and weekend data, but use split tabes for calcuating means and to make the panel plot using the basic plotting system
```{r} 
head(new_tab_df)

str(new_tab_df)
new_tab_df$new_date<-as.Date(new_tab_df$new_date)
new_tab_df$weekday<-weekdays(new_tab_df$new_date)

weekend<-new_tab_df[new_tab_df[,"weekday"]=="Saturday"|new_tab_df[,"weekday"]=="Sunday",]
weekend$day_type<-"weekend"

week<-new_tab_df[new_tab_df[,"weekday"]!="Saturday"|new_tab_df[,"weekday"]!="Sunday",]
week$day_type<-"weekday"

new_with_daytype<-rbind(weekend,week)

mean_steps_week<-tapply(week$steps,week$interval,mean)
mean_steps_week_df<-data.frame(mean_steps_week)
mean_steps_week_df$interval<-unique(week$interval)

mean_steps_weekend<-tapply(weekend$steps,weekend$interval,mean)
mean_steps_weekend_df<-data.frame(mean_steps_weekend)

mean_steps_weekend_df$interval<-unique(weekend$interval)

par(mfrow=c(2,1), oma=c(1,1,1,1))
plot(mean_steps_week~interval, data=mean_steps_week_df, type="l",  xaxt="n",
     ylim=c(0,200), main=c("Weekdays"),xlab=c("Time of day [h]"),ylab=c("Number of steps"))
axis(1, at=c(0, 400, 800, 1200, 1600, 2000,2355),labels=c("00:00","4:00","8:00",
                                                          "12:00","16:00","20:00",
                                                          "23:55"))

plot(mean_steps_weekend~interval, data=mean_steps_weekend_df, type="l", xaxt="n",
     ylim=c(0,200), main=c("Weekends"),xlab=c("Time of day [h]"),
     ylab=c("Number of steps"))
axis(1, at=c(0, 400, 800, 1200, 1600, 2000,2355),labels=c("00:00","4:00","8:00",
                                                          "12:00","16:00","20:00",
                                                          "23:55"))

```

